A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-17 response/response.tex

%%%% First Reviewer %%%%
\section{Re: Information to the authors}
%
\subsection*{Larger issues}
%
\begin{enumerate}
    \item
         Some parts of the manuscript suggest that the authors introduced a new method,
         i.e., the complex-step (CS) method. For example, in the introduction, they say
         “A goal of this study was to develop and evaluate a new, accurate, and practical
         method ...” Another example appears in page 3, where the authors say “The aim
         of this paper is not only to introduce the new complex-step method in the context
         of evaluating tangent-stiffness matrices ..”. However, the CS method has already
         appeared in many publications since 1998 [1]. Please revise the way you refer to
         the CS method along the manuscript

{\color{red}
  \begin{itemize}
      \item
        The way the CS method is referred to has been revised throughout,
        specifically: 
    \item In the abstract, "... the new method ..." has been
        replaced by "... the method ...", "... newly implemented ..." has been
        replaced by "... implemented ..."; 
    \item in the introduction, paragraph two,
        "... a new, accurate ..." has been replaced by "... a relatively
        underexplored, accurate ...", in paragraph two and four "... the new
        method ..." has been replaced by "... the method ...", in paragraph
        five, "... introduce the new complex-step method in the context of ..."
        has been replaced by "... study the complex-step method of ...", "...
        the new complex-step ..." has been replaced by "... the complex-step
        ...".
    \item 
        It is the authors' opinion that the above changes correct the suggestion
        that the authors introduce a new method.

  \end{itemize}}

    \item
        The work presented focuses on a parallel computational peridynamic code, and
        uses peridynamic models for mechanics. However, no detailed explanation of the
        equations involved is presented. Although the tangent-stiffness calculations are
        not specialized to Peridigm, the authors should include a short section introducing
        the peridynamic theory and the specific constitutive model used, i.e., the elastic
        peridynamic solid. You may add this to the already included very brief description
        of peridynamics in page 9 (lines 15-26). This will also help to explain the concept
        of the peridynamic horizon in page 10.

{\color{red}
  \begin{itemize}
    \item 
        We have revised the brief description of peridynamics in page 9 to be more complete:
    \item
        Material from silling2007 is interpreted including: the definition of vector states,
        the derivation of conservation laws in terms of state-based peridynamics, the definition
        of the isotropic elastic solid AKA the Linear elastic Peridynamic Solid.
    \item
        It is the authors' opinion that the additional material mentioned above explains the equations involved
        in the peridynamic models for mechanics, used in relation to the peridynamics code
        studied in the work.

  \end{itemize}}
    \item 
        The manuscript seems to be written in a language more familiar to computer scientists. 
        I would encourage the authors to revise certain parts of the manuscript to be
        presented in a language more familiar to the computational mechanics community.

{\color{red}
  \begin{itemize}
    \item 
        We have revised certain parts of the manuscript for clarity, adding additional information for the reader's
        convenience:
    \item
        In subsection 'Implementing the complex-step method in Peridigm', explanation of what is meant by datatype,
        container class, template, scope and instantiation are included since these concepts may be unfamiliar to
        a computational mechanics audience at least as denoted by these terms. 
    \item 
        In subsection 'The Comparative Study', the way message passing computations can slow runtime is explained
        by an analogy to an experiment given as an exercise for the reader. 
    \item 
        In section 'Conclusions', the term byte-copiable is explained. 
    \item 
        It is the authors' opinion that the above revisions to the language of the manuscript allow the intended audience
        to follow the discussion where concepts from computer science are relevant.
  \end{itemize}}

### 2014-02-20 jacobian_calc_CMAME.tex
added solution accuracy results and convergence studies

>>>>

\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}
In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

\subsection{Common code} 
\label{subsec:CommonCode}

The source code for the header-only library is entitled
\emph{NewtonRaphson.hpp}. Note that the library is an interface to functions and
data structures made available in the \emph{Trilinos} library, mentioned
earlier, such that it is required in order for the examples described here
to work. \emph{Trilinos} is freely available for download as of this
writing at: \emph{trilinos.sandia.gov}.  Generally the header-only library
works by taking in the users initial guessed equilibrium solution, target
dependent variable values, system constants, and a function pointer to a
model evaluation function. Solver parameters including an update scaling
factor, tolerance, maximum iterations and probe length are chosen after the
solver is instantiated as a new object. The behavior of the solver is
specialized through a coding feature called inheritance to perform each of
CS, AD and FD.  But, in common, each version of the solver works by
computing the system Jacobian, solving a linear system as in that step of
the Newton Raphson method, computing the residual and checking for
convergence or overrun. Here the linear solution method is LU
decomposition, while the chief convergence criterion is change in residual
between iterations.  Additionally the problem can be partially respecified
to allow for running a series of load steps having progressing target
values, which is a typical mode also seen in \emph{Peridigm}. Applying
boundary conditions incrementally in load steps assists convergence
\cite{rezaiee2010dynamic}.

\subsection{First example}
\label{subsec:ProgOne}
The first example program, \emph{NRexamples.cpp}, solves a simple nonlinear
system describing the intersection of three infinite paraboloids in threespace.
It works by defining force computation functions for each of the models, and
then passing an address called a function pointer corresponding to each of
these methods to a solver object which coordinates solving the problem and
reporting output for the user based on their convergence criteria and other
preferences like step-size and maximum number of iterations. CS, AD and FD are
each represented and the user is free to adjust input settings by specifying
command line arguments as the compiled program instructs, or perhaps create a
shell script that varies parameters for each of the methods in order to perform
a comparative study.

\subsection{Second example and related studies} 
\label{subsec:Validation}
The second program,a \emph{NRComparison.cpp}, follows a case study that appears in
\cite{rezaiee2010dynamic}. The example program departs from the cited
literature by applying each of CS, AD and FD and verifying them against an
analytical model of a nonlinear mechanical system involving a connected beam
and spring. Fig. \ref{fig:TrussSchematic} is a schematic of the beam and spring
system.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8cm]{./figs/truss.png}
  \caption{Beam and spring simulated in example program.}
  \label{fig:TrussSchematic}
\end{figure}

The equations describing this system and results for verification were obtained
in \cite{rezaiee2010dynamic}. They include a force function, Eqn.
\ref{eqn:TrussForce}, and an analytical force derivative or stiffness, Eqn.
\ref{eqn:TrussStiffness}.

\begin{equation} 
    \label{eqn:TrussForce}
    f(D) = .5AE(cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}cos^{2}\phi - 3sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})sin^{2}\phi,
\end{equation}

\begin{equation} 
    \label{eqn:TrussStiffness}
    S_{T}(D) = 1.5AE(cos^{2}\phi)[\frac{D}{L_{0}}cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AEsin^{2}\phi}{L_{0}}.
\end{equation}

Where $\phi$ is a constant, beam angle, $AE$ is rigidity, $D$ is displacement,
$L_{0}$ is original beam length, and $k_{s}$ is the Hooke's constant for the
coil spring attached to the beam. For a verification of each of the methods as
well as a convergence rate study, this truss and spring model was analyzed in
the second program. $f(D)$ is restorative force to balance downward force
loading at the connection point of the beam and the spring, parallel to the
wall. As in the reference, equilibrium solutions for 12 incremental load steps
of 4.448 Newtons were sought. Differing from the reference, Newton's method was
used rather than 'Dynamic Relaxation'. Solver parameters common to all tests
were a limit of 100 iterations for convergence, a residual change tolerance of
$1\times{10^{-9}} \si{\pico\meter}$ and an update multiplier of 1.0. Data
collected included number of iterations to converge summed over all load-steps
and final displacement, D, after the final load-step. Additionally
finite-difference step-size was varied in order to measure the impact on each
of the Jacobian calculation methods. Naturally, varying step size does not
affect AD or the analytically derived Jacobian. The accuracy results for
verifying the Jacobian calculation methods are summarized in Table
\ref{tab:ConvergenceStudy}.

\begin{table*}[!ht]    
  \centering
        \caption{Displacement, D, after load-step 12} \label{tab:Verification}   
        \begin{tabular}{c c c c c}
         \toprule
         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 
        \midrule
        0 & 5.07996 & " & " & "\\
        -4 & 5.07996 & " & " & "\\
        -8 & 5.07996 & " & " & "\\
        -12 & 5.07996 & " & " & " \\
        -16 & 5.07996 & " & " & failed \\
        -20 & 5.07996 & " & " & failed \\
        \bottomrule
    \end{tabular}
\end{table*}

Table \ref{tab:Verification} shows that each of the methods solved the problem
as accurately as one another within the precision displayed on the table. The
exception to this is that the finite difference method failed due to an
arithmetic error in association with extremely small step size, suggesting
subtractive cancellation. However, since there is no accuracy reason for
choosing an extremely small step size, each of the methods can be used to
produce acceptable solutions to the nonlinear truss system. Table
\ref{tab:ConvergenceStudy} shows the total number of iterations taken over all
load-steps for each of the methods.

\begin{table*}[!ht]    
  \centering
        \caption{Summed iterations over all load steps} \label{tab:ConvergenceStudy}   
        \begin{tabular}{c c c c c}
         \toprule
         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 
        \midrule
        0 & 63 & " & 213 & 417\\
        -4 & 63 & " & " & "\\
        -8 & 63& " & " & "\\
        -12 &63& " & " &  65\\
        -16 &63& " & " & fail\\
        -20 &63& " & " & fail\\
        \bottomrule
    \end{tabular}
\end{table*}

Table \ref{tab:ConvergenceStudy} indicates that when using AD or the 'actual'
analytically derived Jacobian, a minimum number of iterations is achieved
corresponding to a maximum convergence rate. CS and FD are similarly affected
by a grossly large step-size, while subtractive cancellation invalidates the
results for FD for very small step-sizes. Inaccuracy for large step-sizes is is
predicted by the FD and CS formulas, which are contingent upon a small
step-size to allow the neglect of higher order 12 terms. The results of this
test show that step-size selection is not unimportant when using FD, but when
properly configured any of the methods may return the same maximum convergence
rate for this problem.  Overall conclusions regarding the methods that may be
kept in mind for problems besides this example are that the accuracy of the
Jacobian should not be confused with the accuracy of the equi- librium
solution, since each of the methods discussed here may be equally capable given
enough iterations. Additionally, while the accuracy of the Jacobian may
strongly affect convergence rates, this relationship was not quantified here.
